---
title: "Nile River Flows"
author: "Jeffrey Arnold"
date: "September 17, 2015"
output: html_document
---

```{r results = 'hide', echo = FALSE, message = FALSE}
knitr::opts_chunk[["set"]](cache =TRUE,
                           fig.path = "figures/nile-",
                           dev = 'pdf',
                           fig.height = 1.66,
                           fig.width = 3.2,
                           autodep = TRUE)
library("rstan")
library("ggplot2")
library("dplyr")
library("tidyr")
library("loo")
library("xtable")
options(mc.cores = parallel::detectCores())

this <- new.env()

tab_path <- "../analysis/tex"
dir.create(tab_path, showWarnings = FALSE)

theme_local <- theme_minimal
```


```{r helper_funs}
omega_summary <- function(x, param = "omega") {
 omega <- drop(rstan::extract(x, param)[[1]])
 nile_df %>%
  mutate(mean = apply(omega, 2, mean),
         median = apply(omega, 2, median),
         sd = apply(omega, 2, sd),
         z = mean / sd,
         p25 = apply(omega, 2, quantile, prob = 0.25),
         p75 = apply(omega, 2, quantile, prob = 0.75),
         p16 = apply(omega, 2, quantile, prob = 0.16),
         p84 = apply(omega, 2, quantile, prob = 0.84),
         p025 = apply(omega, 2, quantile, prob = 0.025),
         p975 = apply(omega, 2, quantile, prob = 0.975)) 
}

omega_plot <- function(.data) {
  ggplot(.data, aes(x = year)) +
    geom_hline(yintercept = 0, colour = "gray") +
    geom_linerange(mapping = aes(ymin = p025, ymax = p975), colour = "gray") +
    geom_point(mapping = aes(y = mean)) +
    theme_local() + 
    theme(panel.grid = element_blank())
}

omega_z_plot <- function(.data) {
  ggplot(.data, aes(x = year, y = z)) +
    geom_point() +
    #geom_ribbon(ymin = -2, ymax = 2, alpha = 0.2) + 
    geom_hline(yintercept = 0) + 
    theme_minimal()
}


kalman_summary <- function(x, param = "kalman") {
  kal <- drop(rstan::extract(x, param)[[1]])
  nile_df %>%
  mutate(mean = apply(kal, 2, mean),
         median = apply(kal, 2, median),
         p25 = apply(kal, 2, quantile, prob = 0.25),
         p75 = apply(kal, 2, quantile, prob = 0.75),
         p16 = apply(kal, 2, quantile, prob = 0.16),
         p84 = apply(kal, 2, quantile, prob = 0.84),
         p025 = apply(kal, 2, quantile, prob = 0.025),
         p975 = apply(kal, 2, quantile, prob = 0.975))    
}

kalman_plot <- function(x) {
  ggplot(x, aes(x = year)) +
    geom_hline(yintercept = 0, colour = "gray") +
    geom_pointrange(mapping = aes(ymin = p025, ymax = p975, y = mean)) +
    theme_local()
}

mu_summary <- function(x, param = "mu") {
  mu <- drop(rstan::extract(x, param)[[1]])
  if (ncol(mu) == (nrow(nile_df) + 1)) {
    mu <- mu[ , 2:ncol(mu)]
  }
  nile_df %>%
  mutate(mean = apply(mu, 2, mean),
         sd = apply(mu, 2, sd),
         median = apply(mu, 2, median),
         p25 = apply(mu, 2, quantile, prob = 0.25),
         p75 = apply(mu, 2, quantile, prob = 0.75),
         p16 = apply(mu, 2, quantile, prob = 0.16),
         p84 = apply(mu, 2, quantile, prob = 0.84),
         p025 = apply(mu, 2, quantile, prob = 0.025),
         p975 = apply(mu, 2, quantile, prob = 0.975))  
}

mu_plot <- function(.data) {
  ggplot(.data, aes(x = year)) + 
    geom_ribbon(mapping = aes(ymin = p16, ymax = p84),
                alpha = 0.3) +
    geom_ribbon(mapping = aes(ymin = p025, ymax = p975),
                alpha = 0.3) +
    geom_point(mapping = aes(y = flow), size = 1) +
    geom_line(mapping = aes(y = mean)) +
    ylab("") +
    xlab("") + 
    theme_local() +
    theme(panel.grid = element_blank())
}

lambda_summary <- function(x, param = "lambda") {
  lambda <- drop(rstan::extract(x, param)[[1]])
  nile_df %>%
  mutate(mean = apply(lambda, 2, mean),
         sd = apply(lambda, 2, sd),
         median = apply(lambda, 2, median),
         p25 = apply(lambda, 2, quantile, prob = 0.25),
         p75 = apply(lambda, 2, quantile, prob = 0.75),
         p16 = apply(lambda, 2, quantile, prob = 0.16),
         p84 = apply(lambda, 2, quantile, prob = 0.84),
         p025 = apply(lambda, 2, quantile, prob = 0.025),
         p975 = apply(lambda, 2, quantile, prob = 0.975))  
}

lambda_plot <- function(.data) {
  ggplot(.data, aes(x = year)) +
    geom_hline(yintercept = 0, colour = "gray") +
    geom_pointrange(mapping = aes(ymin = p025, ymax = p975, y = mean)) +
    theme_local()
}

dlm_err <- function(x, param = "dlm") {
  rstan::extract(x, param)[[1]][ , -1, 5]
}

```

I compare the three cases considered in @HarchaouiLevy-Leduc2010

1. Low noise: $\sigma = 0.05$
2. Medium noise: $\sigma = 0.1$
3. High Noise: $\sigma = 0.5$

These three cases are each sub-sampled to 1,000 points.

The blocks function is included in this project's R package **bsdlm** function `blocks_dj94`:
```{r messages = FALSE}
STAN_DIR <- "../stan/"
SEED <- 6714590

data("Nile", package = "datasets")
nile_df <- data_frame(year = as.numeric(time(Nile)),
                      flow = as.numeric(Nile))

nile_data <- 
  list(y = nile_df[["flow"]],
       miss = as.integer(is.na(nile_df[["flow"]])),
       n = nrow(nile_df),
       m0 = mean(nile_df[["flow"]]),
       C0 = var(nile_df[["flow"]]) * 5,
       s = sd(nile_df[["flow"]]) * 5
       )

```

```{r nile, fig.height = 2.5, fig.width = 6}
ggplot(nile_df, aes(x = year, y = flow)) + 
  geom_point() + 
  geom_vline(xintercept = 1899, colour = "red") + 
  annotate("text", x = 1902, y = 1200, label = "Expected change point", hjust = 0,
           colour = "red") +
  xlab("") +
  theme_local()
```



# Models

## M0: Constant Mean

$$
\begin{aligned}[t]
y_t &\sim N(\mu, \sigma^2) \\
\mu &\sim N(m_0, C_0)
\end{aligned}
$$

```{r m0}
m0_model <- stan_model(file.path(STAN_DIR, "constant_mean.stan"))
m0_ret <- sampling(m0_model,
                   data = nile_data,
                   seed = SEED)

```

```{r m0_mu}
m0_mu_ <- rstan::extract(m0_ret, "mu")[[1]]
m0_mu <-
  nile_df %>%
  mutate(mean = mean(m0_mu_),
         median = median(m0_mu_),
         sd = sd(m0_mu_),
         p25 = quantile(m0_mu_, prob = 0.25),
         p75 = quantile(m0_mu_, prob = 0.75),
         p16 = quantile(m0_mu_, prob = 0.025),
         p84 = quantile(m0_mu_, prob = 0.975),
         p025 = quantile(m0_mu_, prob = 0.025),
         p975 = quantile(m0_mu_, prob = 0.975))

mu_plot(m0_mu)
```



## M1: Dummies Model

$$
\begin{aligned}[t]
y_t &\sim N(\alpha + X \omega, \sigma^2) \\
\omega &\sim N(0, s) \\
\alpha &\sim N(m_0, C_0)
\end{aligned}
$$

```{r m1}
m1_data <- within(nile_data, {
  w <- sd(nile_df[["flow"]]) * 5
  X <- matrix(as.integer(nile_df[["year"]] >= 1899))
  M <- ncol(X) 
})

m1_model <- stan_model(file.path(STAN_DIR, "changepoint_intervention.stan"))
m1_ret <- sampling(m1_model,
                   data = m1_data,
                   seed = SEED)

```

```{r m1_mu}
m1_mu <- mu_summary(m1_ret)
mu_plot(m1_mu)
```


## M2: Local Level

$$
\begin{aligned}[t]
y_t &\sim N(\mu_t, \sigma^2) \\
\mu_t &\sim N(\mu_{t-1}, \sigma^2 \tau^2)
\sigma &\sim C^{+}(0, s) \\
\tau &\sim C^{+}(0, w) \\
\mu_0 &\sim N(m_0, C_0)
\end{aligned}
$$

```{r m2}
m2_data <- within(m1_data, {
  w <- 1
})
m2_model <- stan_model(file.path(STAN_DIR, "local_level.stan"))
m2_ret <- sampling(m2_model,
                   data = m2_data,
                   seed = SEED)
```

```{r m2_mu}
m2_mu <- mu_summary(m2_ret)
mu_plot(m2_mu)
```

```{r m2_omega}
m2_omega <- omega_summary(m2_ret)
omega_plot(m2_omega)
```

```{r m2_z_omega}
omega_z_plot(m2_omega)
```

```{r m2_kalman}
m2_kalman <- kalman_summary(m2_ret)
kalman_plot(m2_kalman)
```

To confirm that the results are reasonable. 
Note that sigma and tau are scale parameters, while `StructTS` returns variances.
```{r}
summary(m2_ret, par = c("sigma", "tau"))
StructTS(Nile, type = c("level"))
```


## M3: Student's $t$

$$
\begin{aligned}[t]
y_t &\sim N(\mu_t, \sigma^2) \\
\mu_t &\sim N(\mu_{t-1}, \sigma^2 \tau^2 \lambda^2)
\sigma &\sim C^{+}(0, s) \\
\tau &\sim C^{+}(0, w) \\
\lambda^2 &\sim \mathrm{InvGamma}(0.5 * \nu, 0.5 * nu) \\
\nu & \sim \mathrm{Gamma}(2, 0.1) \\
\mu_0 &\sim N(m_0, C_0)
\end{aligned}
$$

```{r m3}
m3_data <- within(m1_data, {
  w <- 1 / length(y)
  nu <- 1
})
m3_model <- stan_model(file.path(STAN_DIR, "changepoint_student_t.stan"))
m3_ret <- sampling(m3_model,
                   data = m3_data,
                   seed = SEED)
```

```{r m3_mu}
m3_mu <- mu_summary(m3_ret)
mu_plot(m3_mu)
```

```{r m3_omega}
m3_omega <- omega_summary(m3_ret)
omega_plot(m3_omega)
```

```{r m3_z_omega}
omega_z_plot(m3_omega)
```

```{r m3_kalman}
m3_kalman <- kalman_summary(m3_ret)
kalman_plot(m3_kalman)
```

```{r m3_lambda}
m3_lambda <- lambda_summary(m3_ret)
lambda_plot(m3_lambda)
```


## M4: Double Exponential (Laplace)

$$
\begin{aligned}[t]
y_t &\sim N(\mu_t, \sigma^2) \\
\mu_t &\sim N(\mu_{t-1}, \sigma^2 \tau^2 \lambda^2)
\sigma &\sim C^{+}(0, s) \\
\tau &\sim C^{+}(0, w) \\
\lambda^2 &\sim \mathrm{Exp}(0.5) \\
\mu_0 &\sim N(m_0, C_0)
\end{aligned}
$$

```{r m4}
m4_data <- within(m1_data, {
  w <- 1 / length(y)
  nu <- 1
  miss <- as.integer(is.na(y))
})
m4_model <- stan_model(file.path(STAN_DIR, "changepoint_double_exponential.stan"))
m4_ret <- sampling(m4_model,
                   data = m4_data,
                   seed = SEED)
```

```{r m4_mu}
m4_mu <- mu_summary(m4_ret)
mu_plot(m4_mu)
```

```{r m4_omega}
m4_omega <- omega_summary(m4_ret)
omega_plot(m4_omega)
```

```{r m4_z_omega}
omega_z_plot(m4_omega)
```

```{r m4_kalman}
m4_kalman <- kalman_summary(m4_ret)
kalman_plot(m4_kalman)
```

```{r m4_lambda}
m4_lambda <- lambda_summary(m4_ret)
lambda_plot(m4_lambda)
```




## M5: Horseshoe

$$
\begin{aligned}[t]
y_t &\sim N(\mu_t, \sigma^2) \\
\mu_t &\sim N(\mu_{t-1}, \sigma^2 \tau^2 \lambda^2)
\sigma &\sim C^{+}(0, s) \\
\tau &\sim C^{+}(0, w) \\
\lambda &\sim C^+(0, 1) \\
\mu_0 &\sim N(m_0, C_0)
\end{aligned}
$$

```{r m5}
m5_data <- within(m1_data, {
  w <- 1 / length(y)
  miss <- as.integer(is.na(y))
})
m5_model <- stan_model(file.path(STAN_DIR, "changepoint_horseshoe.stan"))
m5_ret <- sampling(m5_model,
                   data = m5_data,
                   seed = SEED)
```

```{r m5_mu}
m5_mu <- mu_summary(m5_ret)
mu_plot(m5_mu)
```

```{r m5_omega}
m5_omega <- omega_summary(m5_ret)
omega_plot(m5_omega)
```

```{r m5_z_omega}
omega_z_plot(m5_omega)
```

```{r m5_kalman}
m5_kalman <- kalman_summary(m5_ret)
kalman_plot(m5_kalman)
```

```{r m5_lambda}
m5_lambda <- lambda_summary(m5_ret)
lambda_plot(m5_lambda)
```



## M6: Horseshoe+

$$
\begin{aligned}[t]
y_t &\sim N(\mu_t, \sigma^2) \\
\mu_t &\sim N(\mu_{t-1}, \sigma^2 \tau^2 \lambda_t^2)
\sigma &\sim C^{+}(0, s) \\
\tau &\sim C^{+}(0, w) \\
\lambda_t &\sim C^+(0, \eta_t) \\
\eta_t &\sim C^+(0, 1) \\
\mu_0 &\sim N(m_0, C_0)
\end{aligned}
$$

```{r m6}
m6_data <- within(m1_data, {
  w <- 1 / length(y) * 1000
  miss <- as.integer(is.na(y))
})
m6_model <- stan_model(file.path(STAN_DIR, "changepoint_horseshoeplus.stan"))
m6_ret <- sampling(m6_model,
                   data = m6_data,
                   seed = SEED)
```


```{r m6_mu}
m6_mu <- mu_summary(m6_ret)
mu_plot(m6_mu)
```

```{r m6_omega}
m6_omega <- omega_summary(m6_ret)
omega_plot(m6_omega)
```

```{r m6_z_omega}
omega_z_plot(m6_omega)
```

```{r m6_kalman}
m6_kalman <- kalman_summary(m6_ret)
kalman_plot(m6_kalman)
```

```{r m6_lambda}
m6_lambda <- lambda_summary(m6_ret)
lambda_plot(m6_lambda)
```


## Model Comparison

```{r}
waic2df <- function(x) {
  x_waic <- waic(x)
  x_loo <- loo(x)
  bind_cols(as_data_frame(x_waic[! names(x_waic) %in% c("pointwise")]),
            as_data_frame(x_loo[! names(x_loo) %in% c("pointwise", "pareto_k")]))
}

msae <- function(err) {
  mse <- sqrt(mean((apply(err, 2, mean))^2))
  mae <- mean(abs(apply(err, 2, mean)))
  data_frame(mse = mse, mae = mae)
}

waic_loo <- list(mutate(waic2df(rstan::extract(m0_ret, "log_lik")[[1]]),
            model = "\\ModelII{Constant}"),
      mutate(waic2df(rstan::extract(m1_ret, "log_lik")[[1]]),
            model = "\\ModelII{Intervention}"),
      mutate(waic2df(rstan::extract(m2_ret, "log_lik")[[1]]),
            model = "\\ModelII{Normal}"),
      mutate(waic2df(rstan::extract(m3_ret, "log_lik")[[1]]),
            model = "\\ModelII{StudentT}"),
      mutate(waic2df(rstan::extract(m4_ret, "log_lik")[[1]]),
            model = "\\ModelII{Laplace}"),
      mutate(waic2df(rstan::extract(m5_ret, "log_lik")[[1]]),
            model = "\\ModelII{Horseshoe}"),
      mutate(waic2df(rstan::extract(m6_ret, "log_lik")[[1]]),
            model = "\\ModelII{Horseshoe+}")) %>%
  bind_rows()

mse_mae <- list(mutate(msae(rstan::extract(m0_ret, "err")[[1]]),
                    model = "\\ModelII{Constant}"),
              mutate(msae(rstan::extract(m1_ret, "err")[[1]]),
                    model = "\\ModelII{Intervention}"),
              mutate(msae(dlm_err(m2_ret)),
                    model = "\\ModelII{Normal}"),
              mutate(msae(dlm_err(m3_ret)),
                    model = "\\ModelII{StudentT}"),
              mutate(msae(dlm_err(m4_ret)),
                    model = "\\ModelII{Laplace}"),
              mutate(msae(dlm_err(m5_ret)),
                    model = "\\ModelII{Horseshoe}"),
              mutate(msae(dlm_err(m6_ret)),
                    model = "\\ModelII{Horseshoe+}")) %>%
  bind_rows()

model_comp <- left_join(waic_loo, mse_mae, by = "model")

model_comp %>%
  select(model, mse, elpd_waic, elpd_loo) %>%
  rename(`$\\mathrm{RMSE}$` = mse,
         `$\\mathrm{elpd}_{WAIC}$` = elpd_waic,
         `$\\mathrm{elpd}_{loo}$` = elpd_loo)  %>%
  xtable() %>%
  print(sanitize.text = identity,
        floating = FALSE,
        include.rownames = FALSE,
        type = "latex",
        file = file.path(tab_path, "nile-tab_model_comp.tex"))

```

```{r model_comp_mu}
mu_mse_mae <- function(x, mu_true) {
  if (length(dim(x)) == 1) {
    x_mean <- mean(x)
    mse1 <- sqrt(mean((mu_true - x_mean) ^ 2))
    mse2 <- mean(sqrt(sapply(mu_true, function(i) mean((i - x) ^ 2))))
  } else {
    x <- drop(x)[ , -1]
    x_mean <- apply(x, 2, mean)
    mse1 <- sqrt(mean((mu_true - x_mean) ^ 2))
    mse2_ <- numeric(length(mu_true))
    for (i in seq_along(mu_true)) {
      mse2_[i] <- mean((x[ , i] - mu_true[i]) ^ 2)
    }
    mse2 <- mean(sqrt(mse2_))
  }
  data_frame(mse1 = mse1, mse2 = mse2)
}

mu_true <- m1_mu[["mean"]]

model_comp_mu <- list(mutate(mu_mse_mae(rstan::extract(m0_ret, "mu")[[1]], mu_true),
                    model = "\\ModelII{Constant}"),
                mutate(mu_mse_mae(rstan::extract(m2_ret, "mu")[[1]], mu_true),
                    model = "\\ModelII{Normal}"),
                mutate(mu_mse_mae(rstan::extract(m3_ret, "mu")[[1]], mu_true),
                    model = "\\ModelII{StudentT}"),
              mutate(mu_mse_mae(rstan::extract(m4_ret, "mu")[[1]], mu_true),
                    model = "\\ModelII{Laplace}"),
              mutate(mu_mse_mae(rstan::extract(m5_ret, "mu")[[1]], mu_true),
                    model = "\\ModelII{Horseshoe}"),
              mutate(mu_mse_mae(rstan::extract(m6_ret, "mu")[[1]], mu_true),
                    model = "\\ModelII{Horseshoe+}")) %>%
  bind_rows()

ggplot(gather(model_comp_mu, stat, value, - model),
       aes(x = model, y = value)) +
  geom_point() +
  facet_wrap(~ stat, nrow = 2) +
  coord_flip()

model_comp_mu

model_comp_mu %>%
  select(model, mse1, mse2) %>%
  rename(`$\\mathrm{MSE}_1(\\mu)$` = mse1, `$\\mathrm{MSE}_2(\\mu)$` = mse2)  %>%
  xtable() %>%
  print(sanitize.text = identity,
        floating = FALSE,
        include.rownames = FALSE,
        type = "latex",
        file = file.path(tab_path, "nile-tab_mu_model_comp.tex"))
                         
```



## Save Data

```{r}
saveRDS(as.list(this), file = "nile.rds")
```
